---
title: "Bayesian Practical"
author: "Alba Mas and Maria Cobo"
date: "2023-11-17"
output:
  pdf_document: default
  html_document: default
---

\usepackage[T1,T2A]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[russian,english]{babel} % if you're using Russian text

We are analysing the lifetime of a bacterium at a certain temperature, which follows an Exponential distribution with density: x\>0, $\mu$ \> 0 $$\text{f} (x|\mu) = \frac{1}{\mu}e^{-\frac{x}{\mu}}$$

We observe the lifetime of n=8 bacteria (expressed in hours):

2.56 0.42 3.72 0.64 10.7 1.57 9.62 2.54**\
\
1- Write the likelihood function. Estimate the mean life expectancy** $\mu$ **using a classical inference procedure.**

$$L(\mu) = \text{П} \frac{1}{\mu}e^{-\frac{x}{\mu}} = (\frac{1}{\mu})^ne^{-\frac{1}{\mu}\sum x_i}$$

$$\alpha(\frac{1}{\mu})^8e^{-\frac{31.77}{\mu}}$$

$$ln(\text{L}(\mu|X)) = ln(\frac{1}{\mu})^8 ln(e^{-\frac{31.77}{\mu}}) = -8ln(\mu) - ({\frac{31.77}{\mu}})$$

$$\frac{d}{d\mu}(-8ln(\mu) - ({\frac{31.77}{\mu}}))=\frac{-8}{\mu}-\frac{31.77}{\mu²}=0$$

$$\mu = 3.97$$

**2- The authors of a paper claim that the mean life expectancy of this bacterium is 4.5±0.5h (reading the paper we see that 0.5h is just the standard deviation). Using a Gaussian density for implementing the prior information for parameter** $\mu$**, compute the posterior density, the expectation and a 95% credibility interval. You can follow the following steps:**

**a) Plot the approximated posterior, that is, plot** $\pi$**(**$\mu$**)𝐿(𝑋\|**$\mu$**). This is just the posterior without the normalizing constant.**

```{r}
x = seq(3, 7, 0.1)
prior <- dnorm(x, 4.5, 0.5)

# Likelihood function
likelihood <- function(mu) {
  (1/mu^8) * exp(-31.77/mu)
}

approx_post <- likelihood(x) * prior
plot(x, approx_post, type='l', col='blue', xlab='μ', ylab='π(μ) L(X|μ)', main='Approximated Posterior')
```

**b) Note that this is not a density, you would need that** $\int_{0}^{\infty} \text{L}(\mu|X) \, d\mu = 1$**. Then, construct a proper density dividing the approximated posterior by its integral. You can use the function integrate() in R. Plot the posterior density**

```{r}
approx_post <- function(mu) {
  likelihood(mu) * dnorm(mu, 4.5, 0.5)
}

# Constant
C <- integrate(approx_post, lower = 0, upper = Inf, abs.tol = 1e-09)$value

# Posterior density function
posterior_density <- function(mu) {
  approx_post(mu) / C
}

# Plot the Posterior density
plot(x, posterior_density(x), type = 'l', col = 'hotpink', xlab = 'μ', ylab = 'Posterior Density', main = 'Posterior Density')


```

**c) Plot the posterior distribution function, that is, F(**$\mu$**)=** $\int_{0}^{\mu} \text{f}(\mu|X) \, d\mu$

```{r}
posterior_distribution_function <- function(u) {
  integrate(posterior_density, lower = 0, upper = u)$value
}

# CDF (cumulative distribution function) values for the given range
cdf_values <- sapply(x, posterior_distribution_function)

# Plot posterior distribution function 
plot(x, cdf_values, type = 'l', col = 'blue', xlab = 'μ', ylab = 'Posterior Distribution Function', main = 'Posterior Distribution Function')
```

**d) The expectation can be computed as** $\int_{0}^{\infty} \text{f}(\mu|X) \, d\mu$ **and the credibility interval is obtained by solving F(u)=0.025 and F(u)=0.975. You can solve equations numerically by using function uniroot() in R.**

```{r}
approx_post <- function(mu) {
  likelihood(mu) * dnorm(mu, 4.5, 0.5)
}

integrate(approx_post, 0, Inf)

# Solve F(u) = 0.025
lower <- function(u) integrate(approx_post, 0, u, abs.tol = 1e-09)$value/C - 0.025

# Solve F(u) = 0.975
upper <- function(u) integrate(approx_post, 0, u, abs.tol = 1e-09)$value/C - 0.975

low_CI <- uniroot(lower, c(2, 4.5))$root
low_CI

upp_CI <- uniroot(upper, c(4.5, 6))$root
upp_CI
```

**3- Test if your results agree with those of the authors of the paper. In other words, consider H0: 𝜇 = 4.5 and H1: 𝜇 ≠ 4.5. Be non informative for the prior probabilities of H0 and H1, and consider for H1 the same Gaussian prior than in part 2. Compute P(H0\|X), P(H1\|X), and the Bayes factor. Which are the conclusions?**

To test the hypotheses H0: 𝜇 = 4.5 and H1: 𝜇 ≠ 4.5, we can use Bayes' theorem to calculate the posterior probabilities of each hypothesis given the observed data. The posterior probabilities are given by:

$$P(H0|X) = \frac{P(X|H0)P(H0)}{P(X)}$$

$$P(H1|X) = \frac{P(X|H1)P(H1)}{P(X)}$$

The Bayes factor is the ratio of the posterior odds:

$$Bayes factor = \frac{P(H0|X)}{P(H1|X)}$$

Since the prior probabilities are non-informative we can assume equal prior probabilities P(H0)=P(H1)=0.5

```{r}
prior_H0 <- 0.5
prior_H1 <- 0.5

# Likelihoods
likelihood_H0 <- dnorm(4.5, 4.5, 0.5) 
likelihood_H1 <- integrate(approx_post, lower = 0, upper = Inf, abs.tol = 1e-09)$value / C 

marginal_likelihood <- prior_H0 * likelihood_H0 + prior_H1 * likelihood_H1

posterior_H0 <- (likelihood_H0 * prior_H0) / marginal_likelihood
posterior_H1 <- (likelihood_H1 * prior_H1) / marginal_likelihood

bayes_factor <- posterior_H0 / posterior_H1

cat("Posterior probability of H0:", posterior_H0, "\n")
cat("Posterior probability of H1:", posterior_H1, "\n")
cat("Bayes factor:", bayes_factor, "\n")
```

This results show that the posterior probability of H1​ (mean life expectancy different from 4.5 hours) is higher than the posterior probability of ​H0 (mean life expectancy exactly 4.5 hours). But since the Bayes factor is so low (0.7978846) we don't have very strong evidence in favor of H1​ relative to H0​.
